Actuator line modeling of vertical-axis turbines

### 2016-05-04 paper.tex
Add reference to RVAT baseline dataset
Add more ref to experiments, and always call RM2 medium solidity

Both the high solidity UNH-RVAT and medium-low solidity RM2 turbines were
modeled using the ALM inside a Reynolds-averaged Navier--Stokes (RANS)
simulation, closed with the standard $k$--$\epsilon$ turbulence model. These
rotors provide diverse parameters, which helped evaluate the robustness of the
ALM. The simulations were performed inside a domain similar in size to that used
in \cite{Bachant2016-BR-CFD} to mimic the tow tank experiments (3.66 m wide,
2.44 m tall, 1.52 m upstream and 2.16 m downstream), with similar velocity
boundary conditions: 1 m/s inflow, no-slip (fixed to 1 m/s) walls and bottom,
and a rigid slip condition for the top, to approximate some effects of the tow
tank's free surface. Simulations were run for a total of 6 seconds, with the
latter half used to calculate performance and wake statistics.
Pressure--velocity coupling for the momentum equation was achieved using the
PISO (pressure implicit splitting of operators) method. A slice of the mesh in
the $x$--$y$ plane is shown in Figure~\ref{fig:ALM-mesh}.

@@

    \caption{Power and drag coefficient curves computed for the UNH-RVAT using
        the actuator line model with RANS.}

@@

Figure~\ref{fig:RVAT-ALM-meancontquiv} shows mean velocity field for the
UNH-RVAT computed by the ALM RANS model. The asymmetry was captured well, along
with some of the vertical flow due to blade tip vortex shedding, though the flow
structure is missing the detail present in the experiments and blade-resolved
RANS simulations. Overall, the wake appears to be over-diffused, which could be
a consequence of the relatively coarse mesh. Note that with the DS and flow
curvature corrections turned off, the direction of the mean swirling motion
reverses, which highlights the importance of resolving the correct azimuthal
location of blade loading.

@@

Turbulence kinetic energy contours (including resolved and modeled energy) are
shown in Figure~\ref{fig:RVAT-ALM-kcont}. The ALM was able to resolve the
concentrated area of $k$ on the $+y$ side of the turbine, but the turbulence
generated by the dynamic stall vortex shedding process is absent. This makes
sense since in the ALM, the DS model only modulates the body force term in the
momentum equation, which does not provide a mechanism for mimicking shed
vortices or turbulence.

@@

    \caption{Mean streamwise velocity (left) and turbulence kinetic energy
        (right) profiles at $z/H=0$ for the UNH-RVAT ALM.}

@@

Figure~\ref{fig:RM2-ALM-perf-curves} shows the performance curves computed for
the RM2 by the ALM, and those from the tow tank experiments. As with the high
solidity RVAT, $C_P$ is overpredicted at high $\lambda$. However, $\lambda_0$,
the tip speed ratio of peak power coefficient, is also shifted to the right.
This is indicative of inaccurate dynamic stall modeling, which could possibly be
attributed to one of the models' tuning constants, e.g., the time constant for
the lagged angle of attack. Limited ad hoc testing revealed that $C_P$ at
$\lambda_0$ was more accurately predicted with a lagged angle of attack time
constant roughly double the default value. This could be investigated further by
looking at the phase of, e.g., maximum $C_P$ peaks compared to the experimental
values (or possibly those from blade-resolved CFD), but for the present study
only mean values were considered.

@@

    \caption{Power and drag coefficient curves computed for the RM2 using the
        ALM.}

@@

Figure~\ref{fig:RM2-ALM-meancontquiv} shows the mean velocity field at 1 m
downstream or $x/D=0.93$ computed by the ALM for the RM2. The mean near-wake
structure looks similar to the RVAT ALM RANS case, but for the RM2, the lack of
detail from blade tip vortex shedding matches more closely with experiments.
However, the vertical flow towards the $x$--$y$ center plane was captured, which
is an important qualitative feature of both CFT near-wakes.

@@

    \caption{Mean streamwise velocity (left) and turbulence kinetic energy
    (right) profiles at $x/D=0.93$ and $z/H=0$ computed for the RM2 using the
    ALM inside a $k$--$\epsilon$ RANS model.}

@@

The near-wake's mean velocity field at $x/D=1$ is shown in
Figure~\ref{RVAT-ALM-LES-vorticity}. Compared with the RANS ALM results, the LES
looks much more like the blade-resolved and experimental results, showing the
clockwise and counterclockwise mean swirling motion on the $-y$ and $+y$ sides
of the rotor, respectively.

@@

    \caption{Mean velocity profiles in the UNH-RVAT near-wake at $x/D=1$ and
        $z/H=0$ computed with the Smagorinsky LES model.}

@@

An actuator line model for cross-flow turbines, including a Leishman--Beddoes
type dynamic stall model, flow curvature, added mass, and lifting-line based end
effects corrections, was developed and validated against experimental datasets
acquired for high and medium-low solidity rotors at scales where the performance
and near-wake dynamics were essentially Reynolds number independent. When
coupled to a $k$--$\epsilon$ RANS solver ALM simulations took $O(0.1)$ CPU hours
per second of simulated time, while when coupled with a Smagorinsky LES model
the computing time was $O(10)$ hours per second, which represent a four and two
order of magnitude decrease in computational expense versus 3-D blade-resolved
RANS, respectively.

>>>>

Both the high solidity UNH-RVAT and medium solidity RM2 turbines were modeled
using the ALM inside a Reynolds-averaged Navier--Stokes (RANS) simulation,
closed with the standard $k$--$\epsilon$ turbulence model. These rotors provide
diverse parameters, which helped evaluate the robustness of the ALM. The
simulations were performed inside a domain similar in size to that used in
\cite{Bachant2016-BR-CFD} to mimic the tow tank experiments (3.66 m wide, 2.44 m
tall, 1.52 m upstream and 2.16 m downstream), with similar velocity boundary
conditions: 1 m/s inflow, no-slip (fixed to 1 m/s) walls and bottom, and a rigid
slip condition for the top, to approximate some effects of the tow tank's free
surface. Simulations were run for a total of 6 seconds, with the latter half
used to calculate performance and wake statistics. Pressure--velocity coupling
for the momentum equation was achieved using the PISO (pressure implicit
splitting of operators) method. A slice of the mesh in the $x$--$y$ plane is
shown in Figure~\ref{fig:ALM-mesh}.

@@

    \caption{Power and drag coefficient curves computed for the UNH-RVAT using
        the actuator line model with RANS, compared with experimental results
        from~\cite{Bachant2016-RVAT-Re-dep}.}

@@

Figure~\ref{fig:RVAT-ALM-meancontquiv} shows mean velocity field for the
UNH-RVAT computed by the ALM RANS model. The asymmetry observed in the
experiments \cite{Bachant2015-JoT} was captured well, along with some of the
vertical flow due to blade tip vortex shedding, though the flow structure is
missing the detail present in the experiments and blade-resolved RANS
simulations. Overall, the wake appears to be over-diffused, which could be a
consequence of the relatively coarse mesh. Note that with the DS and flow
curvature corrections turned off, the direction of the mean swirling motion
reverses, which highlights the importance of resolving the correct azimuthal
location of blade loading.

@@

Turbulence kinetic energy contours (including resolved and modeled energy) are
shown in Figure~\ref{fig:RVAT-ALM-kcont}. The ALM was able to resolve the
concentrated area of $k$ on the $+y$ side of the turbine, present in the
experiments~\cite{Bachant2015-JoT}, but the turbulence generated by the dynamic
stall vortex shedding process is absent. This makes sense since in the ALM, the
DS model only modulates the body force term in the momentum equation, which does
not provide a mechanism for mimicking shed vortices or turbulence.

@@

    \caption{Mean streamwise velocity (left) and turbulence kinetic energy
        (right) profiles at $z/H=0$ for the UNH-RVAT ALM, compared with the
        experimental data from~\cite{Bachant2016-RVAT-Re-dep}.}

@@

Figure~\ref{fig:RM2-ALM-perf-curves} shows the performance curves computed for
the RM2 by the ALM, and those from the tow tank
experiments~\cite{Bachant2016-RM2-paper}. As with the high solidity RVAT, $C_P$
is overpredicted at high $\lambda$. However, $\lambda_0$, the tip speed ratio of
peak power coefficient, is also shifted to the right. This is indicative of
inaccurate dynamic stall modeling, which could possibly be attributed to one of
the models' tuning constants, e.g., the time constant for the lagged angle of
attack. Limited ad hoc testing revealed that $C_P$ at $\lambda_0$ was more
accurately predicted with a lagged angle of attack time constant roughly double
the default value. This could be investigated further by looking at the phase
of, e.g., maximum $C_P$ peaks compared to the experimental values (or possibly
those from blade-resolved CFD), but for the present study only mean values were
considered.

@@

    \caption{Power and drag coefficient curves computed for the RM2 using the
        ALM, compared with experimental data from~\cite{Bachant2016-RM2-data}.}

@@

Figure~\ref{fig:RM2-ALM-meancontquiv} shows the mean velocity field at 1 m
downstream or $x/D=0.93$ computed by the ALM for the RM2. The mean near-wake
structure looks similar to the RVAT ALM RANS case, but for the RM2, the lack of
detail from blade tip vortex shedding matches more closely with
experiments~\cite{Bachant2016-RM2-paper}. However, the vertical flow towards the
$x$--$y$ center plane was captured, which is an important qualitative feature of
both CFT near-wakes.

@@

    \caption{Mean streamwise velocity (left) and turbulence kinetic energy
        (right) profiles at $x/D=0.93$ and $z/H=0$ computed for the RM2 using the
        ALM inside a $k$--$\epsilon$ RANS model, compared with experimental data
        from~\cite{Bachant2016-RM2-data}.}

@@

The near-wake's mean velocity field at $x/D=1$ is shown in
Figure~\ref{RVAT-ALM-LES-vorticity}. Compared with the RANS ALM results, the LES
looks much more like the blade-resolved and experimental results
\cite{Bachant2015-JoT}, showing the clockwise and counterclockwise mean swirling
motion on the $-y$ and $+y$ sides of the rotor, respectively.

@@

    \caption{Mean velocity profiles in the UNH-RVAT near-wake at $x/D=1$ and
        $z/H=0$ computed with the Smagorinsky LES model, compared with experimental
        data from~\cite{Bachant2016-RVAT-Re-dep}.}

@@

An actuator line model for cross-flow turbines, including a Leishman--Beddoes
type dynamic stall model, flow curvature, added mass, and lifting-line based end
effects corrections, was developed and validated against experimental datasets
acquired for high and medium solidity rotors at scales where the performance and
near-wake dynamics were essentially Reynolds number independent. When coupled to
a $k$--$\epsilon$ RANS solver ALM simulations took $O(0.1)$ CPU hours per second
of simulated time, while when coupled with a Smagorinsky LES model the computing
time was $O(10)$ hours per second, which represent a four and two order of
magnitude decrease in computational expense versus 3-D blade-resolved RANS,
respectively.
